The yaq Project: \\ A Standard for Scientific Instrumentation

### 2022-10-18 paper.tex
rework sections 3-5

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.

A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
At the bottom, the daemons lay a solid foundation to build the variety of clients on top of.
Client sophistication can be introduced naturally, as novel experiments are tested and refined.

@@

Several general purpose \yaq{} clients exist.
\texttt{yaqc}\cite{yaqc} is a lightweight Python client which is excellent for using in scripts or any other Python program.
\texttt{yaqc} is totally generic and aims to support any conceivable \yaq{} daemon.
\texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
The same client is used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
Technical debt grows especially fast in academic environments where graduate students are constantly being replaced.

We have found that the \yaq{} approach favors many small single-purpose applications above large monolithic ones.
For daemons, the purpose of each application is obvious and unambiguous. 
There is a strict, well defined interface which explicitly limits the kinds of interactions that are provided to the hardware, thus limiting opportunity for unintended consequences.
The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
Tools like \texttt{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing\cite{yaqd-fakes}.
Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.

@@

yaq is also being used in several smaller ways throughout UW-Madison Chemistry.
Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.
This level of collaboration is new to us in the orchestration software space.

>>>>

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
Developing experimental software is an iterative process tied to the scientific goals of the instrument.
Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.

@@

A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
\texttt{yaqc}\cite{yaqc} is a Python client which is excellent for using in scripts or any other Python program.
The code shown creates three client objects which could be used directly through an interactive Python prompt or in a reusable script.
Each client object provides Python methods for each Avro message specified by the associated protocol.

In the middle, the same three hardware peripherals are represented in an interactive, graphical form.
\texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
The self describing \yaq{} interface built using protocols and traits is used to provide interface elements for the most commonly used messages.
This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.

At the top, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Bluesky is a framework which allows users to implement their own experimental procedures.
Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a bluesky procedure and a representation of the resultant data.
\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

All three types of clients represented in Figure \ref{fig:foundation} have been shown addressing the same three hardware peripherals.
All types of clients can be used simultaneously to interact with the same instrument in different modes.
Client sophistication can be introduced naturally, as novel experiments are tested and refined.
Different clients can specialize for different requirements of a custom instrument
For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{Mukamel/Shaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
The same client is used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
Technical debt grows especially fast in academic environments where graduate students are involved in projects for a limited time.

The \yaq{} approach favors many small single-purpose applications above large monolithic ones.
For daemons, the purpose of each application is obvious and unambiguous. 
There is a strict, well defined interface which explicitly limits the kinds of interactions that are provided to the hardware, thus limiting opportunity for unintended consequences.
The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
Tools like \texttt{yaqd-fakes}\cite{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing.
Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.
Integrations with communities like Bluesky offer powerful features which are actively maintained across many institutions.

@@

\yaq{} is open source software.
Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
ZZ individuals have contributed to the development of \yaq{}.

\yaq{} is also being used in several smaller ways throughout UW-Madison Chemistry.
Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.
This level of collaboration is new to us in the orchestration software space.
